---
title: "Statistical Inference Course Project"
author: "Aiyu Li"
date: "01/16/2021"
output: pdf_document
---

# Synopsis

This is a project for Statistical Inference class. There are two parts in this project. 

+ Part 1: simulation Exercise: investigate the distribution of averages of 40 exponentials and a thousand simulations in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. Set lambda = 0.2 for all of the simulations.

+ Part 2: analyze the ToothGrowth data in the R datasets package; provide a basic summary of the data; Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose and provide the compared conclusions.

# R environment and reproducibility

```{r echo=TRUE}
sessionInfo()

set.seed(2021)

```



# Load required libraries

```{r echo=TRUE}
library(stats)
library(ggplot2)
library(sqldf)
library(dplyr, warn.conflicts = F)

```

# Part 1 (P1) Simulation Exercise

## P1 Objectives

+ Show the sample mean and compare it to the theoretical mean of the distribution.
+ Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.
+ Show that the distribution is approximately normal.

## P1: Setup variables and create simulation dataset
```{r echo=TRUE}
# investigate the distribution of averages of 40 exponentials
n <- 40
# set lambda = 0.2 for all of the simulations
lambda <- 0.2
# need to do a thousand simulations 
nosim <- 1000
# Create simulations dataset
simdata <- matrix(rexp(n*nosim, lambda), nrow = nosim, ncol = n)

```

## P1: Compare sample means to theoretical means of the distribution

```{r echo=TRUE}
mns <- apply(simdata,1,mean)
hist(mns,col="light blue",breaks=50, xlab = "Mean",main = "Distribution of simulated means")
abline(v=1/lambda,col="red",lwd=4)
abline(v=round(mean(mns),3),col="green",lwd=3)
legend('topright', c("theoretical mean","avg sample mean"), col=c("red", "green"), lty=c(1,1), lwd=c(3,3), bty = "n")
comMeans<-paste('sample mean =', round(mean(mns),3), ', theoretical mean=', 1/lambda, sep = "", collapse = NULL)
comMeans

```

Above histogram shows: the sample mean is very closed to theoretical mean of distribution when having enough samples of the exponential distribution.

## P1: Compare sample variance to the theoretical variance of the distribution

```{r echo=TRUE}
vars <- apply(simdata,1,var)
hist(vars,col="pink",breaks=50, xlab = "Variance", main = "Distribution of simulated variances")
abline(v=1/lambda^2,col="red",lwd=3)
abline(v=round(mean(vars),3),col="yellow",lwd=2)
legend('topright', c("theoretical variance","sample variance"), col=c("red", "yellow"), lty=c(1,1), lwd=c(3,3), bty = "n")
comVars<-paste('sample variance =', round(mean(vars),3), ', theoretical variance=', 1/lambda^2, sep = "", collapse = NULL)
comVars

```

Above histogram shows: the sample variance is very closed to theoretical variance of distribution when having enough samples of the exponential distribution.

## P1: Compare the results to a normal distribution

```{r echo=TRUE}
## step 1 draw the histgram 
hist(mns,  prob=T,
     main="Density distributions between sample and normal",
     xlab="Sample means from 1000 simulations")

# step 2 draw simulated sample mean density distribution
lines(density(mns), col="green", lty=1)


# step 3 draw therotical mean desnity distibution
x <- seq(min(mns), max(mns), length=100)
y <- dnorm(x, mean=1/lambda, sd=(1/lambda/sqrt(n)))
lines(x, y, col="red", lty=1)

# step 4 draw both sample mean and theoretical Mean
abline(v=1/lambda, col='red', lwd=2)
abline(v=round(mean(mns),3),col="green",lwd=2)

# step 5 insert legends
legend('topright', c("simulation density", "theoretical density"), 
       lty=c(1,1), col=c("green", "red"), lwd=c(2,2))

```

Above histogram with overlay of sample/normal curves shows: the two distribution curves are very similar and normally distributed.

## P1 Conclusion 

+ The sample mean is very close to the theoretical mean of the distribution.
+ The sample variance is very close to the theoretical variance of the distribution.
+ The sample distribution is approximately normal.

